Applied Temporal Analysis: A Complete Run of the FraCaS Test Suite

### 2021-05-04 paper2020/paper.tex
final iwcs 2021 version

\paragraph{GF}
The first phase of the system, parsing, is taken care of by the 
Grammatical Framework (GF, \citet{ranta_grammatical_2004}), which is a
powerful parser generator for natural languages, based on
type-theoretical abstract grammars.  The present work leverages a
syntactic representation of the FraCaS test suite in GF abstract
syntax~\citep{Ljunglof:2012}. Thanks to this, we skip the parsing
phase and avoid any syntactic ambiguity.

For the purpose of this paper, the important feature of GF
syntax is that it aims at a balance of sufficient abstraction to provide a
semantically-relevant structure, but at the same time it embeds
sufficiently many syntactic features to be able to reconstruct
natural-language text. That is, the parse trees generally satisfy the
homomorphism requirement of \citet{montague_english_1970,montague_proper_1974}, and
we can focus on the translation of syntactic trees to logical forms.
%
Consequently, the system presented
here does not aim at textual natural language understanding, but rather
provides a testable, systematic formal semantics of temporal phenomena.
Example \ref{ex:nobel} shows an abstract syntax tree and its
realisation in English.

@@

\paragraph{Statives}
\textit{A contrario}, if \(P\) is \emph{stative}, then we get a time-interval subsumption property:
\vspace{-1.5ex}
\begin{multline*}
\constant {subsumption}_P :\\ [t_3,t_4] ⊆ [t_1,t_2] → P (t_1,t_2) → P (t_3,t_4)
\end{multline*}
This principle is used to reason about problem (314), below:

@@

\fracasex{280}{
\item	Smith wrote a novel in 1991.
\hyp 	Smith wrote a novel in 1992.
}
Here our interpretation is:
\vspace{2ex}
\[\begin{array}{l}
(∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2)) ∧ \\
(∃y. novel(y) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,y,t_3,t_4)) \\
⟶ ⊥ \\
\end{array}
\]
Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t_1,t_2)$ and
$write(smith,y,t_3,t_4)$, because $x \neq y$, and thus we cannot
deduce equality of the intervals $t_1,t_2$ and $t_3,t_4$. In turn, the
hypothesis cannot be refuted.

@@

The final class of lexemes carrying a temporal-dependent semantics are
verbs taking a proposition as argument, like ``finish'', ``start'',
etc. These verbs modify the temporal context in non-trivial
ways. Consider for example ``finish to ...''. The timespan of the
argument of ``finish'' should end within the timespan of the finishing
action:
\vspace{-1.5ex}
\begin{multline*}
⟦\text{finish to \(s\)}⟧(t_0,t_1) = \\ ∃(t_0',t_1'). t_1' ∈ [t_0,t_1] ∧ ⟦s⟧(t_0',t_1')
\end{multline*}

@@

\providecommand\ncases[1]{{\ensuremath{^{#1}}}}
\begin{table}[hbt]
  \centering
  \small
\begin{tabularx}{\columnwidth}{Xr@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r}
Section      & {\kern -2em} \#FraCaS
                          & This        & FC2         & FC & MINE & Nut  & LP  \\ \hline
Quantifiers  & 75         & .93        & .96         & .96    & .77  & .53  & .93  \\
             &            & \ncases{74}& \ncases{74} &        &      &      &     \\
Plurals      & 33         & .79        & .82         & .76    & .67  & .52  & .73 \\
             &            &            &             &     &   &   & \\
Anaphora     & 28         & .79        & .86         &   -    & -    & -    &  -       \\
Ellipsis     & 52         & .81        & .87         &   -    & -    & -    &  -       \\
Adjectives   & 22         & .95        & .95         & .95    & .68  & .32  & .73 \\
             &            & \ncases{20}&  \ncases{20}&     &   &   &   \\
Comparatives & 31         & .65        & .87         & .56    & .48  & .45  &  -       \\
Temporal     & 75         & .73        &  -          &   -    &   -  &  -   &  -       \\
Verbs        & 8          & .75        & .75         &   -    & -    & -    &  -       \\
Attitudes    & 13         & .85        & .92         & .85    & .77  & .46  & .92  \\ 
             &            &            &             &        &      &      &  \\ \hline
Total        & 337        & .81        & .89         & .83    & .69  & .50  & .85  \\
             &            & \ncases{329}& \ncases{259}& \ncases{174}  & \ncases{181}& \ncases{181}& \ncases{142}
  \end{tabularx}
  \caption{Accuracy of our system compared to others.
    ``This" refers to the approach presented in this paper. When a
    system does not handle the nominal number of test cases (shown in
    the second column), the actual number of test cases attempted is
    shown below the accuracy figure, in smaller font.  ``FC''
    refers to the work of \citet{bernardy_type-theoretical_2017}, and ``FC2'' its followup~\citep{bernardy_wide-coverage_2019}. ``MINE" refers
    to the approach of \citet{Mineshima:2015}, ``NUT" to the CCG
    system that utilises the first-order automated theorem prover
    \textit{nutcracker} \cite{bos:2008}, and ``LP"
    to the system presented by \citet{abzianidze_tableau_2015}. A dash
    indicates that no attempt was made for the section. }
  \label{tab:results}
\end{table}
%Additionally, to provide the reader with a better sense of what the
%complete system is capable and not capable of, we list a number of
%correctly and incorrectly handled examples in \cref{fig:correct-examples} and \cref{fig:incorrect-examples}.
\paragraph{Interaction with anaphora}
\label{sec:interaction-with-anaphora}
One reason explaining the lower performance of our system on some
sections of the testsuite is that our interpretation of time interacts
imperfectly with anaphora and ellipsis. Consider the following example:

\vspace{-1.5ex}
\fracasex{232}
{\item	ITEL won more orders than APCOM did.
\item	APCOM won ten orders.
\hyp 	ITEL won at least eleven orders.
}

@@

\citet{bernardy_wide-coverage_2019} presented a logical system for
handling 8 of the 9 sections of the FraCaS test suite, but excluded
section 7, suggesting that it requires many examples that need an
\textit{ad hoc} treatment. Here, we took up this challenge and have
shown that a system similar to theirs can be extended to cover the
remaining section of the test suite, without considerably decreasing
the performance of the rest of the sections. This is indeed a common
problem with logical approaches, namely the fact that one can have
theoretically motivated implementations of individual phenomena,
e.g. anaphora, ellipsis, quantifiers, temporal reference etc., but
when one tries to put all these together into a unified system, this
proves to be a daunting task. We believe that this paper presents an
exception, and provides a system that can deal with all these
different semantic phenomena under a unified system with very good
results.
%
We use the same combination of a number of well-studied tools: type
theory, parsing using the Grammatical Framework (GF), Monadic Dynamic
Semantics and proof assistant technology (Coq). The system achieves an
accuracy of 0.73 on the Temporal Section and 0.81 overall. The whole
system, including data sets, is available at the following url:
\url{https://github.com/GU-CLASP/FraCoq/tree/iwcs2021}.

@@

% Further cases to be included in an extended FraCaS future suite
% involve examples where the interaction between different tenses needs
% to be captured:
% \fracasex{extra3}{
% 	\item	When the phone rang, John had entered the house.  
% 	\hyp 	John entered the house before the phone rang (YES).
% }

@@

\else
\section*{Acknowledgements}
The research reported in this paper was supported by grant 2014-39 from the
Swedish Research Council, which funds the Centre for Linguistic Theory and
Studies in Probability (CLASP) in the Department of Philosophy, Linguistics,
and Theory of Science at the University of Gothenburg. We are grateful to
our colleagues in CLASP for helpful discussion of some of the ideas presented
here. We also thank anonymous reviewers for their useful comments on an
earlier draft of the paper.
\fi
\bibliographystyle{apalike}

>>>>

\paragraph{GF}
The first phase of the system, parsing, is taken care of by the 
Grammatical Framework (GF, \citet{ranta_grammatical_2004}), which is a
powerful parser generator for natural languages, based on
type-theoretical abstract grammars.  The present work leverages a
syntactic representation of the FraCaS test suite in GF abstract
syntax, in effect a GF FraCaS treebank~\citep{Ljunglof:2012}. Thanks to this, we skip the parsing
phase and avoid any syntactic ambiguity.

For the purpose of this paper, the important feature of GF
syntax is that it aims at a balance of sufficient abstraction to provide a
semantically-relevant structure, but at the same time it embeds
sufficiently many syntactic features to be able to reconstruct
natural-language text. That is, the parse trees generally satisfy the
homomorphism requirement of \citet{montague_english_1970,montague_proper_1974}, and
we can focus on the translation of syntactic trees to logical forms.
%
Consequently, the system presented
here does not aim at textual natural language understanding, but rather
provides a testable, systematic formal semantics of temporal phenomena.
Example \ref{ex:nobel} shows an example abstract syntax tree and its
realisation in English.

@@

\paragraph{Statives}
\textit{A contrario}, if \(P\) is \emph{stative}, then we get a time-interval subsumption property:
\vspace{-1.5ex}
\begin{multline*}
\constant {subsumption}_P :\\ [t_3,t_4] ⊆ [t_1,t_2] → P (t_1,t_2) → P (t_3,t_4)
\end{multline*}
This principle is used to reason about problem (314), below (note that ``Smith'' is used as a surname in the FraCaS and can take both feminine and masculine values):

@@

\fracasex{280}{
\item	Smith wrote a novel in 1991.
\hyp 	Smith wrote a novel in 1992.
}
Here our interpretation is:

\[\begin{array}{l}
(∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2)) ∧ \\
(∃y. novel(y) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,y,t_3,t_4)) \\
⟶ ⊥ \\
\end{array}
\]
Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t_1,t_2)$ and
$write(smith,y,t_3,t_4)$, because $x \neq y$, and thus we cannot
deduce equality of the intervals $t_1,t_2$ and $t_3,t_4$. In turn, the
hypothesis cannot be refuted.

@@

The final class of lexemes carrying a temporal-dependent semantics are
verbs taking a proposition as argument, like ``finish'', ``start'',
etc. These verbs modify the temporal context in non-trivial
ways. Consider for example ``finish to ...''. The timespan of the
argument of ``finish'' should end within the timespan of the finishing
action:

\begin{multline*}
⟦\text{finish to \(s\)}⟧(t_0,t_1) = \\ ∃(t_0',t_1'). t_1' ∈ [t_0,t_1] ∧ ⟦s⟧(t_0',t_1')
\end{multline*}

@@

\providecommand\ncases[1]{{\ensuremath{^{#1}}}}
\begin{table}[hbt]
  \centering
  \small
\begin{tabularx}{\columnwidth}{Xr@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r}
Section      & {\kern -2em} \#FraCaS
                          & This        & FC2         & FC & MINE & Nut  & LP  \\ \hline
Quantifiers  & 75         & .93        & .96         & .96    & .77  & .53  & .93  \\
             &            & \ncases{74}& \ncases{74} &        &      &      &     \ncases{44} \\
Plurals      & 33         & .79        & .82         & .76    & .67  & .52  & .73 \\
             &            &            &             &     &   &   & \ncases{24} \\
Anaphora     & 28         & .79        & .86         &   -    & -    & -    &  -       \\
Ellipsis     & 52         & .81        & .87         &   -    & -    & -    &  -       \\
Adjectives   & 22         & .95        & .95         & .95    & .68  & .32  & .73 \\
             &            & \ncases{20}&  \ncases{20}&     &   &   &  \ncases{12} \\
Comparatives & 31         & .65        & .87         & .56    & .48  & .45  &  -       \\
Temporal     & 75         & .73        &  -          &   -    &   -  &  -   &  -       \\
Verbs        & 8          & .75        & .75         &   -    & -    & -    &  -       \\
Attitudes    & 13         & .85        & .92         & .85    & .77  & .46  & .92  \\ 
             &            &            &             &        &      &      & \ncases {9}  \\ \hline
Total        & 337        & .81        & .89         & .83    & .69  & .50  & .85  \\
             &            & \ncases{329}& \ncases{259}& \ncases{174}  & \ncases{174}& \ncases{174}& \ncases{89}
  \end{tabularx}
  \caption{Accuracy of our system compared to others.
    ``This" refers to the approach presented in this paper. When a
    system does not handle the nominal number of test cases (shown in
    the second column), the actual number of test cases attempted is
    shown below the accuracy figure, in smaller font.  ``FC''
    refers to the work of \citet{bernardy_type-theoretical_2017}, and ``FC2'' its followup~\citep{bernardy_wide-coverage_2019}. ``MINE" refers
    to the approach of \citet{Mineshima:2015}, ``NUT" to the CCG
    system that utilises the first-order automated theorem prover
    \textit{nutcracker} \cite{bos:2008}, and ``LP"
    to the system presented by \citet{abzianidze_tableau_2015}. A dash
    indicates that no attempt was made for the section. }
  \label{tab:results}
\end{table}
%Additionally, to provide the reader with a better sense of what the
%complete system is capable and not capable of, we list a number of
%correctly and incorrectly handled examples in \cref{fig:correct-examples} and \cref{fig:incorrect-examples}.
\paragraph{Interaction with anaphora}
\label{sec:interaction-with-anaphora}
One reason explaining the lower performance of our system on some
sections of the testsuite is that our interpretation of time interacts
imperfectly with anaphora and ellipsis. Consider the following example:

\fracasex{232}
{\item	ITEL won more orders than APCOM did.
\item	APCOM won ten orders.
\hyp 	ITEL won at least eleven orders.
}

@@

\citet{bernardy_wide-coverage_2019} presented a logical system for
handling 8 of the 9 sections of the FraCaS test suite, but excluded
section 7, suggesting that it requires many examples that need an
\textit{ad hoc} treatment. Here, we took up this challenge and have
shown that a system similar to theirs can be extended to cover the
remaining section of the test suite, without considerably decreasing
the performance of the rest of the sections. This is indeed a common
problem with logical approaches, namely the fact that one can have
theoretically motivated implementations of individual phenomena,
e.g. anaphora, ellipsis, quantifiers, temporal reference etc., but
when one tries to put all these together into a unified system, this
proves to be a daunting task. We believe that this paper presents an
exception, and provides a system that can deal with all these
different semantic phenomena under a unified system with very good
results.
%
We use the same combination of a number of well-studied tools as \citet{bernardy_wide-coverage_2019} : type
theory, parsing using the Grammatical Framework (GF), Monadic Dynamic
Semantics and proof assistant technology (Coq). The system achieves an
accuracy of 0.73 on the Temporal Section and 0.81 overall. The whole
system, including data sets, is available at the following url:
\url{https://github.com/GU-CLASP/FraCoq/tree/iwcs2021}.

@@

% Further cases to be included in an extended FraCaS future suite
% involve examples where the interaction between different tenses needs
% to be captured:
% \fracasex{extra3}{
% 	\item	When the phone rang, John had entered the house.  
% 	\hyp 	John entered the house before the phone rang (YES).
% }
Further cases to be included in an extended FraCaS future suite
involve examples where the interaction between different tenses needs
to be captured:\footnote{While this work was completed, the work by \cite{vashishtha-etal-2020} was published. The authors present a five datasets to be used for the training of neural models' ability to capture temporal reasoning. It would be interesting to check the amount of data covered, most specifiaclly the level of fine-grainedness of temporal reasoning needed to capture those examples, as compared to what we have been discussing in this paper. We thank an anonymous reviewer for bringing this work to our attention.}

\fracasex{extra3}{
	\item	When the phone rang, John had entered the house.  
	\hyp 	John entered the house before the phone rang (YES).
}

@@

\else
\section*{Acknowledgements}
The research reported in this paper was supported by grant 2014-39 from the
Swedish Research Council, which funds the Centre for Linguistic Theory and
Studies in Probability (CLASP) in the Department of Philosophy, Linguistics,
and Theory of Science at the University of Gothenburg. We are grateful to
our colleagues in CLASP for helpful discussion of some of the ideas presented
here. We also thank anonymous reviewers for their useful comments on an
earlier draft of the paper.\fi

\bibliographystyle{acl_natbib}
